Select-based random access \\ to variable-byte encodings

### 2020-04-15 gradu/msc_english_contents.tex
Updates

A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
reorganizing and a fast $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
has been widely adopted \citep[e.g.][]{Kon17}.

@@

\section{Rank and Select implementation}
\label{chap:implementations}

@@

\begin{table}
\centering
\caption{Memory requirements of SDSL rank and select data structures\label{table:supportsize}}
\begin{tabular}{l||c} 
Structure & Extra size taken\\ 
\hline \hline 
Rank   & \text{25\% of bit array} \\
Rankv5 & \text{6.25\% of bit array}\\
Select & \text{~6-15\% of bit array (TODO: check this}\\
\hline
\end{tabular}
\end{table}

\section{DAC via rank}
Directly addressable VB codes were first introduced by \citep{Bri09}. In their solution, the chunks are stored in separate arrays by their significance. For each integer, the block 
of the least significant chunk is stored in the first array $A_1$, and its continuation bit to the first bit array $B_1$. Then if the number was stored in multiple chunks, the block of the next chunk is
stored to $A_2$ and its continuation bit to $B_2$ and so on. After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure.

@@

\chapter{DAC with select query}

Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
in $B$ and a single $select_1$ query gets the desired starting point, while the forementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
consists of just a few bit operations.

@@

The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64kB+64kB L1 cache, 512kB L2 cache, 6144kB L3 cache and 32GB of DDR3 1333MHz.
The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64). The codes were compiled with \texttt{g++-7 -std=c++11 -DNDEBUG -O3} and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.

\section{VB decoding comparison}
Six different VB decoding algorithms were compared. The Bris-algorithms are from \citep{Bri09}. All algorithms are implemented using data structures and functions from the SDSL 
library \citep{gbmp2014sea}. Each implementation is compressing and decoding uint64\_t type integers. The $rank$ and $select$ implementations are explained in \autoref{chap:implementations}.

\begin{itemize}
  \item Bris4 - Bris implementation with block size 4, using $rank_v$
  \item Bris4v5 - Bris implementation with block size 4, using $rank_{v5}$
  \item Bris8 - Bris implementation with block size 8, using $rank_v$
  \item Bris8v5 - Bris implementation with block size 8, using $rank_{v5}$
  \item Select4 - Proposed implementation with block size 4, using $select_{MCL}$
  \item Select8 - Proposed implementation with block size 8, using $select_{MCL}$
\end{itemize}

Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).

@@

The v5 version of rank naturally has worse performance because more calculations are needed in the last phase of calculating the rank sum. As a trade-off, it requires 
significantly less memory, as seen on Figure~\ref{table:supportsize}.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
    xlabel={average integer byte length},
    ylabel={ms per million queries},
    xmin=1, xmax=8,
    ymin=0, ymax=800,
    xtick={1,2,3,4,5,6,7,8},
    ytick={0,100,200,300,400,500,600,700,800},
    legend pos=north west,
    ymajorgrids=true,
    grid style=dashed,
    width=13cm, height=8cm
]

@@

\section{Memory usage}
Table~\ref{table:results8bit} and Table~\ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures with data sets consisting
 of 50 million 64-bit numbers. The compressed size column has a sum of continuation bit array and the data blocks. For $rank$ and $select$ methods used in this thesis, the total 
space taken by the extra data structures is only a small fraction of the overall size.

@@

Figure~\ref{graph:subarray} shows results for subarray fetching. The data sets used contained mostly 4-bit integers and a few 32-bit integers randomly among them. The number in the x-axis corresponds to the 
number of large integers within 1000 integers and the y-axis has the average time of fetching 50 subsequent numbers from an initially random location.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
    xlabel={number of 32bit integers per 1000 integers},
    ylabel={ms per million subarray queries},
    xmin=1, xmax=16,
    ymin=400, ymax=600,
    xtick={1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16},
    ytick={300,400,500,600,700,800,900,100},
    legend pos=north west,
    ymajorgrids=true,
    grid style=dashed,
    width=13cm, height=8cm
]

@@

\end{axis}
\end{tikzpicture}
\caption{Rank and select performance with subarray fetching} \label{graph:subarray}
\end{figure}

@@

 - add subarray examples / visualization

 - write more about subarray access in results

>>>>

A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
reorganizing and a fast $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
has been widely adopted \citep[e.g.][]{Kon17, Sha16}.

@@

\section{Rank and Select implementation}

@@

\begin{table}
\centering
\caption{Memory requirements of SDSL rank and select data structures\label{table:supportsize}}
\begin{tabular}{l||c} 
Structure & Extra size taken\\ 
\hline \hline 
$Rank_v$   & \text{25\% of bit array} \\
$Rank_{v5}$ & \text{6.25\% of bit array}\\
$Select$ & \text{~6-15\% of bit array (TODO: check this}\\
\hline
\end{tabular}
\end{table}

\section{DAC via rank}
\label{chap:implementations}
Directly addressable VB codes were first introduced by \citep{Bri09}. In their solution, the chunks are stored in separate arrays by their significance. For each integer, the block 
of the least significant chunk is stored in the first array $A_1$, and its continuation bit to the first bit array $B_1$. Then if the number was stored in multiple chunks, the block of the next chunk is
stored to $A_2$ and its continuation bit to $B_2$ and so on. After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure.

@@

\chapter{DAC with select query}
\label{chap:select_impl}
Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
in $B$ and a single $select_1$ query gets the desired starting point, while the forementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
consists of just a few bit operations.

@@

The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64kB+64kB L1 cache, 512kB L2 cache, 6144kB L3 cache and 32GB of DDR3 1333MHz.
The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64) with minimal programs running in the background. The codes were compiled with \texttt{g++-7 -std=c++11 -DNDEBUG -O3} 
and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.

\section{VB decoding comparison}
Six different VB decoding implementations were compared:
\begin{itemize}
  \item Bris4 - Bris implementation with block size 4, using $rank_v$
  \item Bris4v5 - Bris implementation with block size 4, using $rank_{v5}$
  \item Bris8 - Bris implementation with block size 8, using $rank_v$
  \item Bris8v5 - Bris implementation with block size 8, using $rank_{v5}$
  \item Select4 - Proposed implementation with block size 4, using $select_{MCL}$
  \item Select8 - Proposed implementation with block size 8, using $select_{MCL}$
\end{itemize}

The Bris-algorithms are implementations based on \citep{Bri09} using two different $rank$ data structures from the SDSL-library \citep{gbmp2014sea}. The $select$-algorithm is the one proposed in 
this thesis and it is based on the $select_{MCL}$ data structure from the SDSL-library. Two different block sizes were used to illustrate how block size affects the runtime and memory requirement. Sizes 
8 and 4 were chosen because they are relatively easy to implement.

Each implementation is compressing and dompressing 64-bit unsigned integers. The $rank$ and $select$ implementations are explained in detail in \autoref{chap:implementations} 
and \autoref{chap:select_impl}, respectively.

\pagebreak
Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).

@@

The $rank_{v5}$ naturally has worse performance because more calculations are needed in the last phase of calculating the rank sum. As a trade-off, it requires 
significantly less memory than $rank_v$, as seen on Figure~\ref{table:supportsize}.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
    xlabel={average integer byte length},
    ylabel={ms per million queries},
    xmin=1, xmax=8,
    ymin=0, ymax=800,
    xtick={1,2,3,4,5,6,7,8},
    ytick={0,100,200,300,400,500,600,700,800},
    legend pos=north west,
    ymajorgrids=true,
    grid style=dashed,
    width=13cm, height=8.5cm
]

@@

\section{Memory usage}
Tables~\ref{table:results8bit} and~\ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures with data sets consisting
 of 50 million 64-bit numbers. The compressed size column has a sum of continuation bit array and the data blocks. For $rank$ and $select$ methods used in this thesis, the total 
space taken by the extra data structures is only a small fraction of the overall size.

@@

The subarray access experiments were run similarly to the previous experiments. The data sets used contained 50 million numbers each. Most of the numbers were 4-bit long, with a varying amount of 32-bit 
integers randomly among them. The numbers in the x-axis correspond to the number of 32-bit integers per 1000 integers. One million indexes were randomized and both the index numbers and the data sets were 
preloaded into memory. A subarray of length 50 was decompressed from each randomized index location. Some precaution was used in the index randomizing to prevent reading from an out of bounds location.

Figure~\ref{graph:subarray} shows results for subarray fetching. The probability of having a long integer within the subarray range is visible from the results of the $rank$ based methods. When the density of the long integers increase, so does the need to access multiple 
block layers and thus the average time increases. The results of the $select$ based methods visualize the static need of singular $select$ query. The 4-bit version of $select$ requires a few more instructions 
during the block location phase and therefore is a little bit slower overall.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
    xlabel={number of 32bit integers per 1000 integers},
    ylabel={ms per million subarray queries},
    xmin=1, xmax=16,
    ymin=400, ymax=600,
    xtick={1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16},
    ytick={300,400,500,600,700},
    legend pos=north west,
    ymajorgrids=true,
    grid style=dashed,
    width=13cm, height=8cm
]

@@

\end{axis}
\end{tikzpicture}
\caption{$Rank$ and $select$ performance with subarray fetching} \label{graph:subarray}
\end{figure}

@@
